---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(rsoundfont)
```

# rsoundfont

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
<!-- badges: end -->

`{rsoundfont}` is a package for parsing [soundfont](https://en.wikipedia.org/wiki/SoundFont) files which contain audio data representing multiple musical instruments.

Soundfont files are a common way of distributing a set of music instruments as audio samples, and 
are often used for [MIDI](https://en.wikipedia.org/wiki/MIDI) playback support on various computers and mobile devices.

In addition to a structured way for storing multi-instrument data in a single file,
the Soundfont format also includes looping information for each sample which allows
for a sample playtime to be extended by repeating pre-determined sections of the audio.  This 
is useful when simulating (for example) a piano note held for a number of seconds, even though
the actual sample is of a much shorter period.


## What's in the box

* `read_sf2(sf2_filename)` read a SoundFont file into a nested list of R objects.
    * This parses the meta-information about the samples only.
    * Audio data is only read later, as required, when calling `create_sample()`
* `create_instruments(sf)` to create a set of zone generators and modulators for each instrument.
    * This will probably become one of the main methods of post-processing the
      raw results from `read_sf2()`
    * Currently it parses out the zone generators for each instrument which gives 
      information such as:
        * which key range this zone is valid for
        * root key and tuning information
        * which sample should be played for this instrument in this zone.
    * Eventually this becomes the main pathway to playing an instrument at a particular
      note - as if a key had been pressed on a MIDI keyboard.
* `create_sample(sf, inst, dur)` create a playable sample of the given instrument audio.
  Sample will be looped to the given duration. 
    * This will return an `audioSample` object compatible with the
    [`{audio}`](https://cran.r-project.org/package=audio) package.
    * Once created, you can play the sample with `audio::play()`
    * Note: instruments have multiple audio samples.  Choosing which sample should
      be played for a particular key is part of the zone generator information
      returned by `create_instruments()`

## SoundFont Technical Specification

The technical specification for SoundFont v2.04 is available
[online](https://www.synthfont.com/sfspec24.pdf)

This package currently uses the chunk naming conventions from the specification.


## Lazy loading of sound data

Audio data for samples can be large - quality SoundFont files can easily contain
data for hundreds of instruments with a gigbyte of audio samples.  Many of these instruments
will never be used in a particular session.


So rather than loading all this audio data up-front, 
a reference to the soundfont file is kept, and audio data is loaded from the
file as instruments areneeded e.g. in `create_sample()`


## Limitations


* This package should load most SoundFont files with version v2.00 or greater
* This package currently uses the chunk naming conventions from the specification.
  This is not always very easy to read/parse, and users are directed to the 
  specification for interpretation of the data structure.
* Unsupported for now as I couldn't find any test files
    * 24bit Samples (If you know of any 24bit Soundfont files, please let me 
      know in the github issues)
    * Chunks may have odd lengths - in which case a padding byte should be 
      added so that subsequent chunks are word-aligned.   I haven't seen
      any of these in the wild, and this current package will throw an
      error if this is the case.   Please let me know in the github issues
      if this occurs.
    * Stereo samples may be possible in a soundfont, but I haven't seen any 
      just yet.
* No support for `*.sfArk` or `*.sfz` compressed files.
    * Is there an open source convertor/decompressor for these?
      
** Please let me know of any `*.sf2` files which this package cannot parse **



## Future possibilities

### Pitch-shifting & time-stretching

Audio samples within a SoundFont are supplied at one (or maybe a few) different
pitches.  In order to be general useful as an instrument for MIDI playback,
some facility will have to be added for pitch-shifting.



## Installation

You can install from [GitHub](https://github.com/coolbutuseless/rsoundfont) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/rsoundfont')
```

## Example

1. Obtain a soundfont.
    * This example uses `weedgm4_update` SoundFont from [https://www.philscomputerlab.com/general-midi-and-soundfonts.html](https://www.philscomputerlab.com/general-midi-and-soundfonts.html)
2. Load the SoundFont data with `read_sf2()`
3. Examine the soundfont object
4. Create and play samples from the Soundfont

```{r example}
library(rsoundfont)

# Source: https://www.philscomputerlab.com/general-midi-and-soundfonts.html
filename <- '/Users/mike/projectsdata/soundfonts/weedsgm4_update.sf2'

sf <- read_sf2(filename)

# Names of all instruments
head(names(sf$pdta$shdr), 20)

# Some Sample/looping information about an instrument
sf$pdta$shdr[['Mandolin Trem E5']]
```


```{r eval = FALSE}
samp <- create_sample(sf, 'Mandolin Trem E5', 1)
audio::play(samp)
```

<a href="https://raw.githubusercontent.com/coolbutuseless/rsoundfont/main/man/samples/01.wav"> Wave file </a>

```{r eval = FALSE}
samp <- create_sample(sf, 'Mandolin Trem C5', 1)
audio::play(samp)
```

<a href="https://raw.githubusercontent.com/coolbutuseless/rsoundfont/main/man/samples/02.wav"> Wave file </a>


```{r eval = FALSE}
samp <- create_sample(sf, 'Bagpipe Drone 3', 1)
audio::play(samp)
```


<a href="https://raw.githubusercontent.com/coolbutuseless/rsoundfont/main/man/samples/03.wav"> Wave file </a>


## Acknowledgements

* R Core for developing and maintaining the language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository
