---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(rsoundfont)
```

# rsoundfont

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
<!-- badges: end -->

`{rsoundfont}` is a package for parsing [soundfont](https://en.wikipedia.org/wiki/SoundFont) files which contain audio data representing multiple musical instruments.

Soundfont files are a common way of distributing a set of music instruments as audio samples, and 
are often used for [MIDI](https://en.wikipedia.org/wiki/MIDI) playback support on various computers and mobile devices.

In addition to a structured way for storing multi-instrument data in a single file,
the Soundfont format also includes looping information for each sample which allows
for a sample playtime to be extended by repeating pre-determined sections of the audio.  This 
is useful when simulating (for example) a piano note held for a number of seconds, even though
the actual sample is of a much shorter period.

## What's in the box

* `read_sf2(sf2_filename)` read a SoundFont file into a nested list of R objects
* `create_sample(sf, inst, dur)` create a playable sample of the given instrument.
  Sample will be looped to the given duration. 
    * This will return an `audioSample` object compatible with the
    [`{audio}`](https://cran.r-project.org/package=audio) package.
    * Once created, you can play the sample with `audio::play()`

## SoundFont Technical Specification

The technical specification for SoundFont v2.04 is available
[online](https://www.synthfont.com/sfspec24.pdf)

This package currently uses the chunk naming conventions from the specification.

## Limitations


* This package should load most SoundFont files with version v2.00 or greater
* This package currently uses the chunk naming conventions from the specification.
  This is not always very easy to read/parse, and users are directed to the 
  specification for interpretation of the data structure.
* Unsupported for now as I couldn't find any test files
    * 24bit Samples (If you know of any 24bit Soundfont files, please let me 
      know in the github issues)
    * Chunks may have odd lengths - in which case a padding byte should be 
      added so that subsequent chunks are word-aligned.   I haven't seen
      any of these in the wild, and this current package will throw an
      error if this is the case.   Please let me know in the github issues
      if this occurs.
* No support for `*.sfArk` or `*.sfz` compressed files.      
      
** Please let me know of any `*.sf2` files which this package cannot parse **


## Future possibilities

### Pitch-shifting & time-stretching

Audio samples within a SoundFont are supplied at one (or maybe a few) different
pitches.  In order to be general useful as an instrument for MIDI playback,
some facility will have to be added for pitch-shifting.

#### Lazy loading of SoundFont audio data

Audio data for samples can be large - quality SoundFont files can easily contain
a gigbyte of audio samples.  Rather than loading all these up-front, it
should be possible to keep a reference to the file and lazily load audio data
when an instrument is needed.

Currently I am only working with smaller SoundFont files and do not presently 
have the need for this.  Contact me if this is something you need.


## Installation

You can install from [GitHub](https://github.com/coolbutuseless/rsoundfont) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/rsoundfont')
```

## Example

1. Obtain a soundfont.
    * This example uses `weedgm4_update` SoundFont from [https://www.philscomputerlab.com/general-midi-and-soundfonts.html](https://www.philscomputerlab.com/general-midi-and-soundfonts.html)
2. Load the SoundFont data with `read_sf2()`
3. Examine the soundfont object
4. Create and play samples from the Soundfont

```{r example}
library(rsoundfont)

# Source: https://www.philscomputerlab.com/general-midi-and-soundfonts.html
filename <- '/Users/mike/projectsdata/soundfonts/weedsgm4_update.sf2'

sf <- read_sf2(filename)

# Names of all instruments
head(names(sf$pdta$shdr), 20)

# Some Sample/looping information about an instrument
sf$pdta$shdr[['Mandolin Trem E5']]
```


```{r eval = FALSE}
samp <- create_sample(sf, 'Mandolin Trem E5', 1)
audio::play(samp)
```

<a href="https://raw.githubusercontent.com/coolbutuseless/rsoundfont/main/man/samples/01.wav"> Wave file </a>

```{r eval = FALSE}
samp <- create_sample(sf, 'Mandolin Trem C5', 1)
audio::play(samp)
```

<a href="https://raw.githubusercontent.com/coolbutuseless/rsoundfont/main/man/samples/02.wav"> Wave file </a>


```{r eval = FALSE}
samp <- create_sample(sf, 'Bagpipe Drone 3', 1)
audio::play(samp)
```


<a href="https://raw.githubusercontent.com/coolbutuseless/rsoundfont/main/man/samples/03.wav"> Wave file </a>


## Acknowledgements

* R Core for developing and maintaining the language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository
